Follow-up Samples

### 2017-04-11 submission/manuscript.tex

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma (P-value
\textless{} 0.05) but not carcinoma versus advanced adenoma (P-value
\textgreater{} 0.05). There was a trend for increasingly less community
similarity, between samples pre- and post-treatment from adenoma to
advanced adenoma to carcinoma. Despite this, no difference was found in
the relative abundance of any specific OTU before and after treatment
for adenoma, advanced adenoma, or carcinoma groups (P-value
\textgreater{} 0.05). Using Random Forest models to assess whether
changes in post-treatment samples were towards a normal community, only
those with carcinoma had a significant decrease in positive probability
(P-value \textless{} 0.05); providing further evidence that treatment
has the greatest effect in those with carcinoma. The adenoma model used
a total of 62 OTUs, the SRN model used a total of 61 OTUs, and the
carcinoma model used a total of 59 OTUs. A total of 26 OTUs were common
to all three models with many classifying to commensal bacteria (e.g.
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Anaerostipes},
\emph{Blautia}, and \emph{Dorea}). Both chemotherapy and radiation did
not provide any additional changes to the bacterial community in those
treated for carcinoma (Pvalue \textgreater{} 0.05).

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether the microbiome in patients with adenoma (n = 22),
advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
differences between pre- or post-treatment. We found that carcinoma
patients had a more dissimilar bacterial community between their initial
and follow up sample than those with adenoma (P-value \textless{} 0.001)
{[}Figure 1A{]}. Although no significant differences were observed
between advanced adenoma and carcinoma there was an increase in the
dissimilarity of the samples pre- and post-treatment from adenoma (0.55
± 0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78
± 0.15) {[}Figure 1A{]}. The bacterial community structure before and
after surgery was visualized using NMDS for adenoma {[}Figure 1B{]}
(PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
(PERMANOVA \textless{} 0.05). Interestingly, when pre- and
post-treatment samples were compared, regardless of whether they were
adenoma or carcinoma, there was no significant overall difference in
beta diversity (PERMANOVA \textgreater{} 0.05). There was no difference
between pre- and post-treatment samples when investigating alpha
diversity metrics for adenoma, advanced adenoma, or carcinoma {[}Table
S1{]}. Additionally, there was also no difference in the relative
abundance of any specific OTU between pre- and post-treatment samples
for adenoma, advanced adenoma, or carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model:}} Using normal and adenoma individauls from
a separate cohort we created an adenoma model to help assess whether
those with adenoma in this cohort changed towards normal. The range of
model AUC's from 100 runs of 20 repeated 10-fold cross-validation was
0.62 - 0.72 with the AUC of the model used for classification having an
AUC of 0.65 {[}Figure S2A{]}. There was a total of 62 OTUs in this model
with the vast majority classifying to bacteria typically thought of as
commensal {[}Figure S3A{]}. There was a significant difference between
the actual and predicted disease classification (P-value \textless{}
0.05). There was also no significant decrease in the positive
probability of adenoma between pre- and post-treatment samples (P-value
\textgreater{} 0.05) {[}Figure 2A{]}.

\textbf{\emph{Advanced Adenoma Model:}} Using normal and advanced
adenoma individuals from a separate cohort we created an advanced
adenoma model to help assess whether those with advanced adenoma in this
cohort changed towards normal. The range of model AUC's from 100 runs of
20 repeated 10 fold cross-validation was 0.68 - 0.77 with the AUC of the
model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
There was a total of 61 OTUs in the advanced adenoma model {[}Figure
S3B{]}. Similar to the adenoma model the vast majority of OTUs
classified to bacteria typically thought of as commensal. Also similar
to the adenoma model there was a significant difference between the
actual and predicted disease classification (P-value \textless{} 0.05)
and no significant decrease in the positive probability of advanced
adenoma between pre- and post-treatment samples (P-value \textgreater{}
0.05) {[}Figure 2B{]}.

\textbf{\emph{Carcinoma Model:}} Using normal and carcinoma individuals
from a separate cohort we created a carcinoma model to help assess
whether those with carcinoma in this cohort changed towards normal. The
range of model AUC's from 100 runs of 20 repeated 10 fold
cross-validation was 0.84 - 0.9 with the AUC of the model used for
classification being 0.88 {[}Figure S2C{]}. Interestingly, the AUCs
improved from adenoma to advanced adenoma to carcinoma {[}Figure 2{]}.
There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
Similar to the adenoma and SRN models the vast majority of OTUs
classified to bacteria typically thought of as commensal but OTUs that
also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas} are also important for carcinoma classification
{[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
there was a significant difference between the actual and predicted
disease classification (P-value \textless{} 0.05). There was a
significant decrease in the positive probability of carcinoma between
pre- and post-treatment samples (P-value \textless{} 0.05) {[}Figure
2C{]}; suggesting that the carcinoma samples changed towards normal
after treatment, unlike either adenoma or advanced adenoma. The one
individual still positive for carcinoma after treatment had an increase
in carcinoma positive proability between their pre- and post-treatment
sample {[}Figure 2C{]}.

\textbf{\emph{Adenoma, Advanced Adenoma, and Carcinoma Common OTUs:}}
Next, to identify which OTUs are important at all stages of disease, we
identified common predictive OTUs within the adenoma, advanced adenoma,
and carcinoma models. When we compared the three different models with
each other there were a total of 26 common OTUs. Some of the most common
taxonomic identifications belonged to \emph{Bacteroides},
\emph{Blautia}, \emph{Anaerostipes}, \emph{Lachnospiraceae}, and
\emph{Dorea}. The vast majority of the OTUs that were common between
these models had classifications to bacteria typically thought of as
commensal {[}Table S2{]}.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differs and how it might be used as an early screening tool
{[}9,10,13--15{]}. Here we show that the microbiome changes towards
normal after treatment for carcinoma and that chemotherapy and radiation
did not provide an additive change. Although some of the important OTUs
classified to genera from bacteria considered the ``usual suspects''"
(e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and \emph{Parvimonas})
many did not. The majority of important OTUs had taxonomic
classifications for resident gut microbes and were common for the
adenoma, advanced adenoma, and carcinoma models. This suggests that
members within the commensal community may be the first that change
during CRC pathogenesis. These subtle changes, in turn, could be the
first step in allowing more inflammatory bacteria to gain a foothold
within the colon {[}11{]}.

Unlike previous studies on the microbiome and CRC, ours focuses on
comparing the microbiome during recovery from treatment in adenoma,
advanced adenoma, and carcinoma groups. Although there were differences
for genera associated with specific bacterium linked with CRC {[}Figure
S3{]}, the majority of important OTUs taxonmically classified to
commensal bacteria {[}Figure S3{]}. Although these changes may be
subtle, due to the lack of significant differences in the bacterial
community pre- and post-treatment in adenoma and advanced adenoma
{[}Figure 1{]}; they support the hypothesis that the first members of
the community to change and potentially stay changed even after
treatment are those that are commensal bacteria.

@@

A limitation, in our study, was that there was a significant difference
in the time elapsed in the collection of the follow up sample between
adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
with time passed being less for adenoma (255 ± 42 days) and advanced
adenoma (250 ± 41) than carcinoma (351 ± 102). These results would
indicate that some of the differences observed between the carcinoma and
adenoma groups could be due to differences in collection time.
Specifically, it could confound the observation that carcinomas changed
more than adenomas {[}Figure 1{]}. However, there are two reasons that
this may not be the case. First, the advanced adenoma group had a higher
dissimilarity than adenoma but lower dissimilarity then carcinoma and
the collection time to their follow up samples was less than the adenoma
group. Second, this confounding would not affect the observations where
models were used since they were built using a different cohort
{[}Figures 2 \& S2-S3{]}.

@@

Assessment of the most important OTUs to the model involved counting the
number of times an OTU was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different splits run. This was then
followed with filtering of this list to variables that were only present
in more than 50\% of these 100 runs. The final collated list of
variables was then run through the mtry optimization again. Once the
ideal mtry was found the entire sample set specific to normal versus
adenoma, normal versus advanced adenoma, or normal versus carcinoma was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. For all three models the final
optimized mtry was
\texttt{r}ifelse(as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} ==
as.data.frame(count(srn\_AUC\_run\_summary, best\_mtry)
\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} \&
as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]} ==
as.data.frame(count(crc\_AUC\_run\_summary, best\_mtry)
\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]},
as.data.frame(count(adn\_AUC\_run\_summary, best\_mtry)
\%\textgreater{}\% slice(which.max(n))){[}, ``best\_mtry''{]}, ``not the
same'')`

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}23{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected.

@@

>>>>

\textbf{Results:} There was a larger change in the bacterial community
in response to treatment for carcinoma versus adenoma (P-value
\textless{} 0.05) but not carcinoma versus advanced adenoma (P-value
\textgreater{} 0.05). There was a trend for increasingly less community
similarity, between samples pre- and post-treatment from adenoma to
advanced adenoma to carcinoma. Despite this, no difference was found in
the relative abundance of any specific OTU before and after treatment
for adenoma, advanced adenoma, or carcinoma groups (P-value
\textgreater{} 0.05). Using Random Forest models to assess whether
changes in post-treatment samples were towards a normal community, only
those with carcinoma had a significant decrease in positive probability
(P-value \textless{} 0.05); providing further evidence that treatment
has the greatest effect in those with carcinoma. The adenoma model used
a total of 62 OTUs, the SRN model used a total of 61 OTUs, and the
carcinoma model used a total of 59 OTUs. A total of 26 OTUs were common
to all three models with many classifying to commensal bacteria (e.g.
\emph{Lachnospiraceae}, \emph{Bacteroides}, \emph{Anaerostipes},
\emph{Blautia}, and \emph{Dorea}). Both chemotherapy and radiation did
not provide any additional changes to the bacterial community in those
treated for carcinoma (P-value \textgreater{} 0.05).

@@

\textbf{\emph{The Bacterial Community:}} Within our 67-person cohort we
tested whether the microbiome in patients with adenoma (n = 22),
advanced adenoma (n = 19), or carcinoma (n = 26) had any broad
differences between pre- and post-treatment samples. We found that
carcinoma patients had a more dissimilar bacterial community between
these samples than those with adenoma (P-value \textless{} 0.001)
{[}Figure 1A{]}. Although no significant differences were observed
between advanced adenoma and carcinoma there was an increase in the
dissimilarity of pre- and post-treatment samples from adenoma (0.55 ±
0.21 (mean ± SD)) to advanced adenoma (0.65 ± 0.25) to carcinoma (0.78 ±
0.15) {[}Figure 1A{]}. The bacterial community structure pre-
post-treatment was visualized using NMDS for adenoma {[}Figure 1B{]}
(PERMANOVA \textgreater{} 0.05), advanced adenoma {[}Figure 1C{]}
(PERMANOVA \textgreater{} 0.05), and carcinoma {[}Figure 1D{]}
(PERMANOVA \textless{} 0.05). Interestingly, when pre- and
post-treatment samples were compared, regardless of whether they were
adenoma or carcinoma, there was no significant overall difference in
beta diversity (PERMANOVA \textgreater{} 0.05). There was no difference
between pre- and post-treatment samples when investigating alpha
diversity metrics for adenoma, advanced adenoma, or carcinoma {[}Table
S1{]}. Additionally, there was also no difference in the relative
abundance of any specific OTU between these samples for adenoma,
advanced adenoma, or carcinoma only {[}Figure S1{]}.

\textbf{\emph{Adenoma Model:}} Using normal and adenoma individuals from
a separate cohort we created an adenoma model to help assess whether
those with adenoma in this cohort changed towards normal. The range of
model AUC's from 100 runs of 20 repeated 10-fold cross-validation was
0.62 - 0.72 with the AUC of the model used for classification having an
AUC of 0.65 {[}Figure S2A{]}. There was a total of 62 OTUs in this model
with the vast majority classifying to bacteria typically thought of as
commensal {[}Figure S3A{]}. There was a significant difference between
the actual and predicted disease classification (P-value \textless{}
0.05). There was also no significant decrease in the positive
probability of adenoma between pre- and post-treatment samples (P-value
\textgreater{} 0.05) {[}Figure 2A{]}.

\textbf{\emph{Advanced Adenoma Model:}} Using normal and advanced
adenoma individuals from a separate cohort we created an advanced
adenoma model to help assess whether those with advanced adenoma in this
cohort changed towards normal. The range of model AUC's from 100 runs of
20 repeated 10-fold cross-validation was 0.68 - 0.77 with the AUC of the
model used for classification having an AUC of 0.73 {[}Figure S2B{]}.
There was a total of 61 OTUs in the advanced adenoma model {[}Figure
S3B{]}. Similar to the adenoma model the vast majority of OTUs
classified to bacteria typically thought of as commensal. Also similar
to the adenoma model there was a significant difference between the
actual and predicted disease classification (P-value \textless{} 0.05)
and no significant decrease in the positive probability of advanced
adenoma between pre- and post-treatment samples (P-value \textgreater{}
0.05) {[}Figure 2B{]}.

\textbf{\emph{Carcinoma Model:}} Using normal and carcinoma individuals
from a separate cohort we created a carcinoma model to help assess
whether those with carcinoma in this cohort changed towards normal. The
range of model AUC's from 100 runs of 20 repeated 10-fold
cross-validation was 0.84 - 0.9 with the AUC of the model used for
classification being 0.88 {[}Figure S2C{]}. Interestingly, the AUCs
improved from adenoma to advanced adenoma to carcinoma {[}Figure 2{]}.
There was a total of 59 OTUs in the carcinoma model {[}Figure S3C{]}.
Similar to the adenoma and advanced adenoma models the vast majority of
OTUs classified to bacteria typically thought of as commensal but OTUs
that also classified to \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas} were also important for carcinoma classification
{[}Figure S3C{]}. Also, like the adenoma and advanced adenoma models
there was a significant difference between the actual and predicted
disease classification (P-value \textless{} 0.05). There was a
significant decrease in the positive probability of carcinoma between
pre- and post-treatment samples (P-value \textless{} 0.05) {[}Figure
2C{]}; suggesting that the carcinoma samples changed towards normal
after treatment, unlike either adenoma or advanced adenoma. The one
individual still positive for carcinoma after treatment had an increase
in carcinoma positive proability between their pre- and post-treatment
sample {[}Figure 2C{]}.

\textbf{\emph{Common OTUs:}} Next, to identify which OTUs were important
at all stages of disease, we identified common predictive OTUs within
the adenoma, advanced adenoma, and carcinoma models. When we compared
the three different models with each other there were a total of 26
common OTUs. Some of the most common taxonomic identifications belonged
to \emph{Bacteroides}, \emph{Blautia}, \emph{Anaerostipes},
\emph{Lachnospiraceae}, and \emph{Dorea}. The vast majority of the OTUs
that were common between these models had classifications to bacteria
typically thought of as commensal {[}Table S2{]}. The rank of these
OTUs, based on median mean decrease in accuracy (MDA), varied higly
depending on whether they were in the adenoma, advanced adenoma, or
carcinoma model {[}Figure 3{]}. As an example, \emph{Bacteroides}
(OTU85) was ranked 14\textsuperscript{th} in the carcinoma model but in
both the adenoma and advanced adenoma model ranked
35\textsuperscript{th} and 41\textsuperscript{st} respectively {[}Figure
3{]}.

@@

This study builds upon previous work from numerous labs that have
considered both how the bacterial community between those with and
without CRC differs and how it might be used as an early screening tool
{[}9,10,13--15{]}. Here we show that the microbiome changes towards
normal after treatment for carcinoma and that chemotherapy and radiation
did not provide any additional change. Although some of the important
OTUs classified to genera from bacteria considered the ``usual
suspects'' (e.g. \emph{Fusobacterium}, \emph{Porphyromonas}, and
\emph{Parvimonas}) many did not. The majority of important OTUs had
taxonomic classifications for resident gut microbes and were common for
the adenoma, advanced adenoma, and carcinoma models. This suggests that
members within the commensal community may be the first that change
during CRC pathogenesis. These subtle changes, in turn, could be the
first step in allowing more inflammatory bacteria to gain a foothold
within the colon {[}11{]}.

Unlike previous studies on the microbiome and CRC, ours focuses on
comparing the microbiome during recovery from treatment in adenoma,
advanced adenoma, and carcinoma groups. Although there were differences
for genera associated with specific bacterium linked with CRC {[}Figure
S3{]}, most of the important OTUs taxonomically classified to commensal
bacteria {[}Figure S3{]}. Although these changes may be subtle, due to
the lack of significant differences in the bacterial community pre- and
post-treatment in adenoma and advanced adenoma {[}Figure 1{]}; they
support the hypothesis that the first members of the community to change
and potentially stay changed even after treatment are those that are
commensal bacteria.

@@

A limitation in our study was that there was a significant difference in
the time elapsed in the collection of the post-treatment sample between
adenoma or advanced adenoma versus carcinoma (P-value \textless{} 0.05),
with time passed being less for adenoma (255 ± 42 days) and advanced
adenoma (250 ± 41) than carcinoma (351 ± 102). These results would
indicate that some of the differences observed between the carcinoma and
adenoma groups could be due to differences in collection time.
Specifically, it could confound the observation that carcinomas changed
more than adenomas {[}Figure 1{]}. However, there are two reasons that
this may not be the case. First, the advanced adenoma group had a higher
dissimilarity than adenoma but lower dissimilarity then carcinoma and
the collection time to their follow up samples was less than the adenoma
group. Second, this confounding would not affect the observations where
models were used since they were built using a different cohort
{[}Figures 2 \& S2-S3{]}.

@@

Assessment of the most important OTUs to the model involved counting the
number of times an OTU was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different splits run. This was then
followed with filtering of this list to variables that were only present
in more than 50\% of these 100 runs. The final collated list of
variables was then run through the mtry optimization again. Once the
ideal mtry was found the entire sample set specific to normal versus
adenoma, normal versus advanced adenoma, or normal versus carcinoma was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. For all three models the final
optimized mtry was 2

@@

\textbf{\emph{Statistical Analysis:}} The R software package (v3.3.2)
was used for all statistical analysis. Comparisons between bacterial
community structure utilized PERMANOVA {[}23{]} in the vegan package
(v2.4.1). Comparisons between probabilities as well as overall OTU
differences between pre- and post-treatment samples utilized a paired
Wilcoxson ranked sum test. Where multiple comparison testing was
appropriate, a Benjamini-Hochberg (BH) correction was applied {[}24{]}
and a corrected P-value of less than 0.05 was considered significant.
Unless otherwise stated the P-values reported are those that were BH
corrected. Model rank importance was determined by obtining the median
MDA from the 100 20 repeated 10-fold cross validation and then ranking
from largest to smallest MDA.

@@

\textbf{Figure 3: Common OTUs Model Rank Based on Median MDA.} For each
common OTU the lowest taxonomic identification and importance rank for
each model run is shown.
